dyngen: a multi-modal simulator for spearheading new single-cell omics analyses

### 2021-03-29 manuscript/v3/render.Rmd
add submission files

---
title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
author:
- Robrecht Cannoodt*
- Wouter Saelens*
- Louise Deconinck
- Yvan Saeys
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    citation_package: biblatex
biblatexoptions: sorting=none,url=false
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \usepackage{pifont}
  \usepackage{float}
  \usepackage{caption}
  \captionsetup[table]{position=bottom}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}

@@

\begin{table}[H]
    \caption{
      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette. $^2$ As showcased by Van den Berge et al.  \cite{vandenberge_trajectorybaseddifferentialexpression_2020}.
    } \label{tab:comparison}
    \small
    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
    \blackline
    & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
    \blackline
    \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
    \blackline
    - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
    - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
    - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
    - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
    - & Reaction activity & \no & \no & \no & \no & \yes \\
    \blackline
    \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
    \blackline
    - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
    - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
    - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
    - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
    - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
    - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
    - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
    - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
    \blackline
    \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
    \blackline
    - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
    - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
    - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
    - & Time-series & \no & \no & \no & \no & \yes \\
    - & Snapshot & \no & \no & \no & \no & \yes \\
    \blackline
    \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
    \blackline
    - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
    - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
    - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
    - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
    - & Differential expression & \yes & \no & \no & \no & \yes $ ^2$ \\ \grayline
    - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
    - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
    - & Trajectory alignment & \no & \no & \no & \no & \yes \\
    \blackline
    %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
    %\blackline
    %- & Number of cells &  &  &  &  & \\ \grayline
    %- & Number of genes &  &  &  &  & \\ \grayline
    %- & Execution time & & & & & \\
    \blackline

\end{tabular}
\end{table}

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{result_files/usecase_trajectory_alignment/usecase.pdf}
    \caption{
    \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
 \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}. In \textbf{C}, the individual mappings of the alignment between cells are shown. In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix. In E the complete count matrices were used to perform the alignment. In F, smoothed pseudocells were used.
  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
    }
    \label{fig:traj_align}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/usecase.pdf}
    \caption{
    \textbf{dyngen allows benchmarking of RNA velocity methods.} 
    \textbf{A:} An example bifurcating cycle dataset, with as illustration the expression and ground truth velocity of a gene D1\_TF1 that goes up and down in one branch of the trajectory. 
\textbf{B:} The RNA velocity estimates of gene D1\_TF1 by the different methods.
\textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the velocity arrow cosine. The velocity correlation is the correlation between the ground-truth velocity (A, right) and the predicted velocity (B). The velocity arrow cosine is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C). 
    }
    \label{fig:velocity}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=.8\linewidth]{result_files/usecase_network_inference/cni.pdf}
    \caption{
    \textbf{dyngen allows benchmarking Cell-specific Network Inference (CSNI) methods.} 
    \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left). However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
    \textbf{B:} CSNI methods were executed to predict the regulatory interactions that are active in each cell specifically. Using the ground-truth cell-specific GRN, the performance of each method was quantified on 14 dyngen datasets. 
    }
    \label{fig:scgrn}
\end{figure}

>>>>

---
title: "dyngen: a multi-modal simulator for spearheading new single-cell omics analyses"
author:
- Robrecht Cannoodt*
- Wouter Saelens*
- Louise Deconinck
- Yvan Saeys
date: "14 September 2020"
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    citation_package: biblatex
biblatexoptions: sorting=none,url=false
classoption:
- table
- 10pt
- a4paper
header-includes: |
  \usepackage{tcolorbox}
  \usepackage{colortbl}
  \usepackage{booktabs}
  \usepackage{tabularx}
  \usepackage{fontspec}
  \usepackage{pifont}
  \usepackage{float}
  \usepackage{caption}
  \captionsetup[table]{position=bottom}
  \setmainfont [Path = fonts/,
    UprightFont = *-300,
    ItalicFont = *-300-Italic,
    BoldFont = *-700,
    BoldItalicFont = *-700-Italic
  ]{MuseoSans}

@@

\begin{table}[H]
    \caption{
      \textbf{Feature comparison of single-cell synthetic data generators and simulators.} $ ^1$ As showcased by vignette. $^2$ As showcased by Van den Berge et al.  \cite{vandenberge_trajectorybaseddifferentialexpression_2020}.
    } \label{tab:comparison}
    \small
    \begin{tabular}{p{.25cm}l|*{5}{>{\centering\arraybackslash}p{1.25cm}|}}
            \blackline
            & \method{} & \method{splatter} & \method{powsimR} & \method{PROSSTT} & \method{SymSim} & \method{dyngen} \\
            \blackline
            \multicolumn{7}{l}{\textbf{Available modality outputs}} \\
            \blackline
            - & mRNA expression & \yes & \yes & \yes & \yes & \yes \\ \grayline
            - & Pre-mRNA expression & \no & \no & \no & \no & \yes \\ \grayline
            - & Protein expression & \no & \no & \no & \no & \yes \\ \grayline
            - & Promotor activity & \no & \no & \no & \yes & \yes \\ \grayline
            - & Reaction activity & \no & \no & \no & \no & \yes \\
            \blackline
            \multicolumn{7}{l}{\textbf{Available ground-truth outputs}} \\
            \blackline
            - & True counts & \yes & \no & \no & \yes & \yes \\ \grayline
            - & Cluster labels & \yes & \yes & \no & \yes & \yes \\ \grayline
            - & Trajectory & \yes & \no & \yes $ ^1$ & \yes & \yes \\ \grayline
            - & Batch labels & \yes & \no & \no & \no & \yes $ ^1$\\ \grayline
            - & Differential expression & \no & \no & \no & \yes & \yes $ ^2$ \\ \grayline
            - & Knocked down regulators & \no & \no & \no & \no & \yes \\ \grayline
            - & Regulatory network & \no & \no & \no & \yes & \yes \\ \grayline
            - & Cell-specific regulatory network & \no & \no & \no & \no & \yes \\
            \blackline
            \multicolumn{7}{l}{\textbf{Emulate experimental effects}} \\
            \blackline
            - & Single-cell RNA sequencing & \yes & \yes & \yes & \yes & \yes \\ \grayline
            - & Batch effects & \yes & \no & \no & \no & \yes $ ^1$ \\ \grayline
            - & Knockdown experiment & \no & \no & \no & \no & \yes \\ \grayline
            - & Time-series & \no & \no & \no & \no & \yes \\ \grayline
            - & Snapshot & \no & \no & \no & \no & \yes \\
            \blackline
            \multicolumn{7}{l}{\textbf{Evaluation applications}} \\
            \blackline
            - & Clustering & \yes & \yes & \no & \yes & \no \\ \grayline
            - & Trajectory inference & \yes & \no & \yes & \yes & \yes \\ \grayline
            - & Network inference & \no & \no & \no & \yes & \yes \\ \grayline
            - & Cell-specific network inference & \no & \no & \no & \no & \yes \\ \grayline
            - & Differential expression & \yes & \no & \no & \no & \yes $ ^2$ \\ \grayline
            - & Batch effect correction & \yes & \no & \no & \no & \yes \\ \grayline
            - & RNA Velocity & \no & \no & \no & \no & \yes \\ \grayline
            - & Trajectory alignment & \no & \no & \no & \no & \yes \\
            \blackline
            %\multicolumn{7}{l}{\textbf{Showcased dataset}} \\
            %\blackline
            %- & Number of cells &  &  &  &  & \\ \grayline
            %- & Number of genes &  &  &  &  & \\ \grayline
            %- & Execution time & & & & & \\
            \blackline
\end{tabular}
\end{table}

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{result_files/usecase_trajectory_alignment/supp_fig.pdf}
    \caption{
            \textbf{dyngen allows benchmarking of trajectory alignment methods.} 
 \textbf{A, B}: Two samples containing a linear trajectory, generated by dyngen.
  \textbf{C, D}: Result of the DTW alignment on the samples in \textbf{A} and \textbf{B}. In \textbf{C}, the individual mappings of the alignment between cells are shown. In \textbf{D}, the accumulated distance matrix between the two trajectories is shown, including the black warping path, corresponding to these cell to cell alignments in \textbf{C}.
  \textbf{E, F}: Shows the accumulated distance matrices obtained after using DTW on two trajectories where noise (noise level of 0.4) was added to the count matrix. In E the complete count matrices were used to perform the alignment. In F, smoothed pseudocells were used.
  \textbf{G:} Shows the influence of added noise to the different processing methods. We can see that DTW + smoothing performs best in noisy circumstances.
    }
    \label{fig:traj_align}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{result_files/usecase_rna_velocity/supp_fig.pdf}
    \caption{
            \textbf{dyngen allows benchmarking of RNA velocity methods.} 
            \textbf{A:} An example bifurcating cycle dataset, with as illustration the expression and ground truth velocity of a gene D1\_TF1 that goes up and down in one branch of the trajectory. 
\textbf{B:} The RNA velocity estimates of gene D1\_TF1 by the different methods.
\textbf{C:} The velocity stream plots produced from the predictions of each method, as generated by scvelo.
\textbf{D:} The predictions scored by two different metrics, the velocity correlation and the velocity arrow cosine. The velocity correlation is the correlation between the ground-truth velocity (A, right) and the predicted velocity (B). The velocity arrow cosine is the cosine similarity between the direction of segments of the ground-truth trajectory (A, left) and the RNA velocity values calculated at those points (C). 
    }
    \label{fig:velocity}
\end{figure}

\begin{figure}[H]
    \centering
    \includegraphics[width=.8\linewidth]{result_files/usecase_network_inference/supp_fig.pdf}
    \caption{
            \textbf{dyngen allows benchmarking Cell-specific Network Inference (CSNI) methods.} 
            \textbf{A:} A cell is simulated using the global gene regulatory network (GRN, top left). However, at any particular state in the simulation, only a fraction of the gene regulatory interactions are active.
            \textbf{B:} CSNI methods were executed to predict the regulatory interactions that are active in each cell specifically. Using the ground-truth cell-specific GRN, the performance of each method was quantified on 14 dyngen datasets. 
    }
    \label{fig:scgrn}
\end{figure}
